### 2015-08-05 whoop.tex
Done with Sec 3

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique Boogie constant. Representing all locks statically, instead of their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
%
Currently, \whoop supports Linux kernel mutex and spinlock operations.  In addition, it is easy to enhance it with knowledge of other locking primitives.
%
If \whoop cannot reliably infer a unique lock instance (e.g., because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning. However, we have never encountered this in practice since a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

@@

To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from~\cite{bardsley2014engineering}.  Suppose we are analyzing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region, $\mathit{MR}$.  When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_{\mathit{MR}}$, representing some unspecified index into $\mathit{MR}$; we call this the \emph{watched offset} for $\mathit{MR}$.  We then attempt to prove that it is impossible for $T$ and $U$ to race on $\mathit{MR}$ at index $\mathit{watched}_{\mathit{MR}}$.  If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $\mathit{MR}$, since the watched offset was arbitrary.  This technique of choosing an arbitrary index to analyze for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $\mathit{MR}$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $\mathit{MR}$.

@@

Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.  The idea is to automatically generate likely program invariants based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.  We give two examples; for clarity we use notation from the simple shared variable concurrent programming model of \S\ref{sec:symbolicpairwise}.  If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as post-conditions for $\instrument{f}{i}$.  These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.  Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be wrong, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.  We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculate candidates that are later deemed to be incorrect, and thus discarded by Houdini.  The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

@@

\noindent\textbf{Verification and Error Reporting }
%
For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, is sent to the Boogie verification engine. For each procedure in the program, Boogie generates a VC and discharges it to the Z3 theorem prover.  In particular, the verification for the root-level procedure, encoding the sequential program sketched in Fig.~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair. Successful verification implies that the pair is race-free, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user. To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code. The following is a race that \whoop found and reported for the example of Fig.~\ref{fig:data_race_example}:

@@

\noindent\textbf{Optimizations }
%
We have implemented various optimizations to increase the precision and performance of \whoop.  We comment on the two most effective optimizations.

First, we enriched \whoop with information regarding \emph{kernel-imposed serialization}, to increase precision. The Linux kernel can serialize calls to entry points, thus forcing them to run in sequence instead of an interleaved manner. For example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock. We discovered this when \whoop reported many races between a number of networking entry points of the r8169 driver (see \S\ref{evaluation}); when we investigated the source of these races, we found out that these entry points could not race in reality because of RTNL. \whoop exploits this knowledge and does not create pairs for entry points that are mutually serialized by the kernel. This is an ongoing manual effort: the more drivers we study, the more such properties we discover to make \whoop more precise. Currently, this information is hard-coded inside \whoop. In the future, we plan to provide a configuration file where the user can denote such kernel-specific information.

@@

As well as inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) SMACK is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses; and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.  However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.

@@

Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time. This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a global flag that indicates device availability: if the flag is set to true, then the checking entry point executes, else it blocks. Because \whoop over-approximates read accesses to shared variables, and thus this flag, it can falsely report a pair as racy. However,~\cite[p.\ 173]{corbet2005linux} advises against serializing drivers in this way, as it ``inhibits user ingenuity'' (e.g.\ a user might expect that a driver can be accessed concurrently for performance).

Statically analyzing drivers requires ``closing'' the environment by abstracting away the low-level implementation details. To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in false positives. However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision. Making our model more precise is an ongoing manual effort, but requires Linux kernel expertise. We argue that further work on the model is orthogonal to the contributions of this paper. Also, even if our symbolic analysis results in false positives, \whoop can still use the results to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.

>>>>

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique Boogie constant. Representing all locks statically, instead of their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
%
Currently, \whoop supports Linux kernel mutex and spinlock operations; it is easy to enhance it with knowledge of other locking primitives.
%
If \whoop cannot reliably infer a unique lock instance (e.g.\ because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning. However, we have never encountered this in practice since a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

@@

To avoid quantifiers and the associated theorem proving burden, we use instead a \emph{watchdog race-checking instrumentation}, adapted from previous work~\cite{bardsley2014engineering}.  Suppose we are analyzing entry points $T$ and $U$, and that after translation into Boogie these entry points share a common memory region $\mathit{MR}$.  When analyzing $T$ and $U$ for races, we introduce an unconstrained symbolic constant $\mathit{watched}_{\mathit{MR}}$, representing some unspecified index into $\mathit{MR}$; we call this the \emph{watched offset} for $\mathit{MR}$.  We then attempt to prove that it is impossible for $T$ and $U$ to race on $\mathit{MR}$ at index $\mathit{watched}_{\mathit{MR}}$.  If we can succeed in proving this, we know that $T$ and $U$ must be race-free for the \emph{whole} of $\mathit{MR}$, since the watched offset was arbitrary.  This technique of choosing an arbitrary index to analyze for each map manipulated by an entry point pair can be seen as a form of quantifier elimination: rather than asking the underlying theorem prover to reason for all indices of $\mathit{MR}$, in a quantified manner, we eliminate the quantifier in our encoding, and instead ask the theorem prover to reason about a single, but arbitrary, index of $\mathit{MR}$.

@@

Houdini does not generate the initial pool of candidates: \whoop generates them using a set of heuristics, and passes them to Houdini as a starting point.  This is done based on syntactic patterns extracted from an inter-procedural pass over the code for an entry point.  We give two examples; for clarity we use notation from the simple concurrent programming model of \S\ref{sec:symbolicpairwise}.  If we observe syntactically that procedure $f$ of entry point $T$ may write to, but does not read from, shared variable $s$, then when instrumenting $T$ with symbol $i$, we guess $s \in \hasbeenwritten{i}$ and $s \notin\hasbeenread{i}$ as post-conditions for $\instrument{f}{i}$.  These guesses may be incorrect, for instance if the potential write to $s$ turns out to be in dead code, or if a read from $s$ has already been issued on entry to $\instrument{f}{i}$.  Similarly, if syntactic analysis indicates that $f$ may unlock mutex $m$, we guess $m \notin \currentlockset{i}$ as a post-condition for $\instrument{f}{i}$; this guess may be incorrect, for instance if the unlock operation is not reachable or if a subsequent lock operation acquires the mutex again.  We stress that guessing incorrect candidate invariants does not compromise the soundness of verification: \whoop is free to speculate candidates that are later deemed to be incorrect, and thus discarded by Houdini.  The balance we try to strike is to have \whoop generate sufficient candidates to enable precise lockset analysis, without generating so many candidates that the speed of the Houdini algorithm is prohibitively slow.

@@

\noindent\textbf{Verification and Error Reporting }
%
For each entry point pair, the instrumented sequential program, equipped with procedure and loop summaries, is given to the Boogie verification engine. For each procedure in the program, Boogie generates a VC and discharges it to the Z3 theorem prover.  In particular, the verification for the root-level procedure, implementing the sequential program sketched in Fig.~\ref{fig:sequentialization}, encodes the race-freedom check for the entry point pair. Successful verification implies that the pair is race-free, while an error (i.e.\ counterexample) denotes a \emph{potential} data race and is reported to the user. To improve usability, \whoop has a built-in error reporter that matches counterexamples to source code. The following is a race that \whoop found and reported for the example of Fig.~\ref{fig:data_race_example}:

@@

\noindent\textbf{Optimizations }
%
We have implemented various optimizations to increase the precision and performance of \whoop.  We comment on the two most effective ones.

First, we enriched \whoop with information regarding \emph{kernel-imposed serialization} to increase precision. The Linux kernel can serialize calls to entry points, thus forcing them to run in sequence instead of an interleaved manner. For example, a large number of networking entry points are mutually serialized with RTNL, a network-specific kernel lock. We discovered this when \whoop reported many races between a number of networking entry points of the r8169 driver (see \S\ref{evaluation}); when we investigated the source of these races, we found out that these entry points could not race in reality because of RTNL. \whoop exploits this knowledge and does not create pairs for entry points that are mutually serialized by the kernel. This is an ongoing manual effort: the more drivers we study, the more such properties we discover to make \whoop more precise.
%Currently, this information is hard-coded inside \whoop. In the future, we plan to provide a configuration file where the user can denote such kernel-specific information.

@@

In addition to inheriting soundness issues arising from currently unknown bugs in \whoop and in the external tools that \whoop relies on, we acknowledge that: (i) SMACK is subject to sources of unsoundness, e.g.\ it models integers as an infinite set (rather than as a finite set of bit-vectors), and its memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses; and (ii) that the combination of Clang and SMACK commits our approach to specific choices related to undefined and implementation-defined aspects of the C language when translating to Boogie.  However, \whoop makes no fundamental assumptions related to these translation decisions, meaning that a more accurate C-to-Boogie translation would automatically lead to a more accurate analysis with \whoop.

@@

Another limitation of \whoop is that it is unable to verify drivers designed to be accessed by only a single process at a time. This \emph{single-open device}~\cite[p.\ 173]{corbet2005linux} mode can be enforced by atomically testing (at the beginning of an entry point) a global flag that indicates device availability: if the flag is set, then the checking entry point executes, else it blocks. Because \whoop over-approximates read accesses to shared variables, and thus this flag, it can falsely report a pair as racy. However, experts~\cite[p.\ 173]{corbet2005linux} advise against serializing drivers in this way, as it ``inhibits user ingenuity'' (e.g.\ a user might expect that a driver can be accessed concurrently for performance).

Statically analyzing drivers requires ``closing'' the environment by abstracting away the low-level implementation details. To this end, we developed a simple model for the Linux kernel that consists of (nondeterministic) stub functions. A limitation of our model is that it can, and will, ultimately result in false positives. However, we currently only focus on finding data races, and thus can get away with over-approximating large parts of the Linux kernel, without losing too much precision. Making our model more precise is an ongoing manual effort that requires Linux kernel expertise. We argue that further work on the model is orthogonal to the contributions of this paper. Also, even if our symbolic analysis results in false positives, \whoop can still use the results to significantly speedup a more precise bug-finder, as seen in \S\ref{corral} and \S\ref{evaluation}.
